A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-09-07 paper/paper.methods.tex
Add hard line breaks for easier editing

\subsection{Sets and Indices} \label{sec:meth-sets-indices}
We denote the set of all services %or I think services is good (vs. services/divisions) as you define two types of services (ID vs. HIV) in the problem statement
that clinicians in a single department can provide as $\mathcal{S}$. The following formulation assumes that all clinicians are able to provide all services. The set of all clinicians in the department is denoted as $\mathcal{C}$. The sets of blocks and weekends that clinicians will be assigned to are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The block size used in our experiments is 2 weeks, but the following LP formulation does not specify a particular size for blocks, and so it can be adapted to the needs of the department in question.
%The size of a block is not constrained and can be adapted %how?
%to the needs of the given department. 
A subset of weekends are denoted $\mathcal{L}$, corresponding to established long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly, we denote with $\mathcal{BR}_c$ and $\mathcal{WR}_c$ the block and weekend requests of clinicians as subsets of all blocks and weekends, respectively. For instance, if clinician $c$'s requests intersect with blocks 1 and 2, and weekend 1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table \ref{tbl:sets-indices} presents a summary of the sets and indices described.

@@

\subsection{Variables} \label{sec:meth-variables}
Since each clinician may be assigned to work on any service, during any block of the year, we denote such an assignment as a binary variable $X_{c, b, s}$. A value of 1 indicates that the given clinician $c$ is assigned to provide service $s$ during block $b$. Weekend assignments are similarly defined using a binary variable $Y_{c, w}$, but without a service index, as clinicians are expected to provide all services during the weekends. We then introduce a set of constants $m_{c, s}$ and $M_{c, s}$ to constrain the minimal and maximal number of blocks each clinician is allowed to work during the year. Table \ref{tbl:variables-constants} presents a summary of the constants and variables in the problem.
%To optimize the soft constraint Block-Weekend Adjacency, we maximize the product $X_{c, b, s} \cdot Y_{c, w}$ for adjacent blocks and weekends. To formulate such an objective as a linear function of variables, we introduce another set of variables, denoted by $Z_{c, b, s}$, with additional constraints on its range. Further details regarding this variable are described in Section \ref{sec:meth-objectives}.

\begin{table}[h]
	\centering
	\begin{tabular}{ c l }
		\hline
		\textbf{Name}              & \textbf{Description}                                               \\ \hline
		$X_{c, b, s} \in \{0, 1\}$ & assignment of clinician $c$ to service $s$ on block $b$            \\
		$Y_{c, w} \in \{0, 1\}$    & assignment of clinician $c$ on weekend $w$                         \\
%		$Z_{c, b, s} \in \{0, 1\}$ & helper variable for optimizing Block-Weekend adjacency             \\
		$m_{c, s}$                 & minimum number of blocks clinician $c$ should cover on service $s$ \\
		$M_{c, s}$                 & maximum number of blocks clinician $c$ should cover on service $s$
	\end{tabular}
	\caption{Description of variables and constants in the problem}
	\label{tbl:variables-constants}
\end{table}

@@

\subsection{Objectives} \label{sec:meth-objectives}
As described in Section \ref{sec:problem}, the soft constraints of the clinician scheduling problem include: satisfying clinician block off requests (BR), satisfying clinician weekend off requests (WR), and assigning weekends closer to blocks (BWA). We convert these soft constraints into linear objective functions of the binary variables defined in Section \ref{sec:meth-variables}. Objectives (\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be written as the following linear functions:  %minor suggestion - consider numbering the objectives for ease later when saying things like "these two objectives"...or writing the full name of the objectives instead of 'this' and 'these' as sometimes can get confusing what the 'this' and the 'these' refer to. 
\begin{align}
	&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} (-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, s} \tag{BR} \label{eqn:obj-block-requests}\\
	&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W} (-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{WR} \label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$ holds and 0 otherwise. In the above two objectives, we penalize any assignments that conflict with a block or weekend request, and aim to maximize the non-conflicting assignments. \\

@@

%So if clinician $c$ is assigned to work during block 3, corresponding to weeks 5 and 6 assuming 2-week blocks, they might also want to be assigned to work during weekend 5. In that case, we would like $X_{c, b=3, s} \cdot Y_{c, w=5}$ to be 1, since that indicates both variables are assigned. If at least one of the two variables is not assigned, the product will be 0. 
% I really liked the example explanation!
%For instance, in the above example we will have $\varphi(b) = 2b - 1$. \\

However, as it is, $Q_3$ is not a linear function of its variables and cannot be optimized in a linear programming framework. An approach used to convert such functions into linear objectives involves introducing a helper variable and additional constraints \cite{hammer_boolean_1968}. In our case, introducing a variable $Z_{c, b, s}$ for every product $X_{c, b, s} \cdot Y_{c, w}$ with $w = \varphi(b)$, and constraining $Z$ such that 
\begin{align}
	&Z_{c, b, s} \leq X_{c, b, s} \label{eqn:helper-x-constraint}\\
	&Z_{c, b, s} \leq Y_{c, w=\varphi(b)} &&\forall s\in\mathcal{S} \label{eqn:helper-y-constraint}
\end{align}
allows us to rewrite $Q_3$ as a linear function of $Z$,
\begin{align}
	&Q_3(Z) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} Z_{c, b, s}
\end{align}
Indeed from Eqns. (\ref{eqn:helper-x-constraint}) and (\ref{eqn:helper-y-constraint}),  whenever $X_{c, b, s} \cdot Y_{c, w} = 1$ (respectively, 0), $Z_{c, b, s}$ can attain a maximum value of 1 (respectively, 0), giving us the correct adjacency maximization objective. \\
%Indeed, whenever $X_{c, b, s} \cdot Y_{c, w} = 1$, $Z_{c, b, s}$ can attain a maximum value of 1, and whenever $X_{c, b, s} \cdot Y_{c, w} = 0$, at least one of $X_{c, b, s}$ or $Y_{c, w}$ must be 0, so $Z_{c, b, s}$ will be constrained to attain a maximum value of 0, giving us the correct adjacency maximization objective. \\

@@

%Thus, our clinician scheduling problem is a multiple objective optimization problem. The most common approach to solving multiple objective optimization problems is by optimizing a weighted sum of the normalized objective functions, as this guarantees the optimal solution to be Pareto optimal [ref \ref{???}]. This is the approach we decided to use in our problem, to ensure all three objectives are considered when finding a solution. Under the assumption that each clinician in $\mathcal{C}$ provides all types of services in $\mathcal{S}$, the normalized objectives can be written as follows,
%\begin{align}
%	&\bar{Q}_1(X) = \frac{Q_1(X)}{C \cdot B \cdot S} \tag{Block Requests} \label{eqn:norm-obj-block-requests}\\
%	&\bar{Q}_2(Y) = \frac{Q_2(Y)}{C \cdot W} \tag{Weekend Requests} \label{eqn:norm-obj-weekend-requests} \\
%	&\bar{Q}_3(Z) = \frac{Q_3(Z)}{C \cdot B \cdot S} \tag{Block-Weekend Adjacency} \label{eqn:norm-obj-block-weekend-adj}
%\end{align}
%where we divide each of the original objective functions by the sum of the absolute values of its coefficients [ref \ref{???}]. The final weighted objective is given by
%\begin{equation}
%	\alpha \bar{Q}_1(X) + \beta \bar{Q}_2(Y) + (1 - \alpha - \beta) \bar{Q}_3(Z)
%\end{equation}
%with $0 \leq \alpha, \beta \leq 1$. \\

>>>>

\subsection{Sets and Indices} \label{sec:meth-sets-indices}
We denote the set of all services %or I think services is good (vs.
%services/divisions) as you define two types of services (ID vs. HIV) in the
%problem statement
that clinicians in a single department can provide as $\mathcal{S}$. The
following formulation assumes that all clinicians are able to provide all
services. The set of all clinicians in the department is denoted as
$\mathcal{C}$. The sets of blocks and weekends that clinicians will be assigned
to are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The block size
used in our experiments is 2 weeks, but the following LP formulation does not
specify a particular size for blocks, and so it can be adapted to the needs of
the department in question.
%The size of a block is not constrained and can be adapted %how?
%to the needs of the given department. 
A subset of weekends are denoted $\mathcal{L}$, corresponding to established
long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly,
we denote with $\mathcal{BR}_c$ and $\mathcal{WR}_c$ the block and weekend
requests of clinicians as subsets of all blocks and weekends, respectively. For
instance, if clinician $c$'s requests intersect with blocks 1 and 2, and weekend
1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table
\ref{tbl:sets-indices} presents a summary of the sets and indices described.

@@

\subsection{Variables} \label{sec:meth-variables}
Since each clinician may be assigned to work on any service, during any block of
the year, we denote such an assignment as a binary variable $X_{c, b, s}$. A
value of 1 indicates that the given clinician $c$ is assigned to provide service
$s$ during block $b$. Weekend assignments are similarly defined using a binary
variable $Y_{c, w}$, but without a service index, as clinicians are expected to
provide all services during the weekends. We then introduce a set of constants
$m_{c, s}$ and $M_{c, s}$ to constrain the minimal and maximal number of blocks
each clinician is allowed to work during the year. Table
\ref{tbl:variables-constants} presents a summary of the constants and variables
in the problem.
%To optimize the soft constraint Block-Weekend Adjacency, we maximize the
%product $X_{c, b, s} \cdot Y_{c, w}$ for adjacent blocks and weekends. To
%formulate such an objective as a linear function of variables, we introduce
%another set of variables, denoted by $Z_{c, b, s}$, with additional constraints
%on its range. Further details regarding this variable are described in Section
%\ref{sec:meth-objectives}.

\begin{table}[h]
	\centering
	\begin{tabular}{ c l }
		\hline
		\textbf{Name}              & \textbf{Description}                             
		\\ \hline
		$X_{c, b, s} \in \{0, 1\}$ & assignment of clinician $c$ to service $s$ on
		block $b$            \\
		$Y_{c, w} \in \{0, 1\}$    & assignment of clinician $c$ on weekend $w$       
		\\
		%		$Z_{c, b, s} \in \{0, 1\}$ & helper variable for optimizing Block-Weekend
		%adjacency             \\
		$m_{c, s}$                 & minimum number of blocks clinician $c$ should
		cover on service $s$ \\
		$M_{c, s}$                 & maximum number of blocks clinician $c$ should
		cover on service $s$
	\end{tabular}
	\caption{Description of variables and constants in the problem}
	\label{tbl:variables-constants}
\end{table}

@@

\subsection{Objectives} \label{sec:meth-objectives}
As described in Section \ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section \ref{sec:meth-variables}. Objectives
(\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be
written as the following linear functions:  %minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
(-1)^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, s} \tag{BR}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
(-1)^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{WR}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments. \\

@@

%So if clinician $c$ is assigned to work during block 3, corresponding to weeks
%5 and 6 assuming 2-week blocks, they might also want to be assigned to work
%during weekend 5. In that case, we would like $X_{c, b=3, s} \cdot Y_{c, w=5}$
%to be 1, since that indicates both variables are assigned. If at least one of
%the two variables is not assigned, the product will be 0. 
% I really liked the example explanation!
%For instance, in the above example we will have $\varphi(b) = 2b - 1$. \\

However, as it is, $Q_3$ is not a linear function of its variables and cannot be
optimized in a linear programming framework. An approach used to convert such
functions into linear objectives involves introducing a helper variable and
additional constraints \cite{hammer_boolean_1968}. In our case, introducing a
variable $Z_{c, b, s}$ for every product $X_{c, b, s} \cdot Y_{c, w}$ with $w =
\varphi(b)$, and constraining $Z$ such that 
\begin{align}
&Z_{c, b, s} \leq X_{c, b, s} \label{eqn:helper-x-constraint}\\
&Z_{c, b, s} \leq Y_{c, w=\varphi(b)} &&\forall s\in\mathcal{S}
\label{eqn:helper-y-constraint}
\end{align}
allows us to rewrite $Q_3$ as a linear function of $Z$,
\begin{align}
&Q_3(Z) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S} Z_{c, b, s}
\end{align}
Indeed from Eqns. (\ref{eqn:helper-x-constraint}) and
(\ref{eqn:helper-y-constraint}),  whenever $X_{c, b, s} \cdot Y_{c, w} = 1$
(respectively, 0), $Z_{c, b, s}$ can attain a maximum value of 1 (respectively,
0), giving us the correct adjacency maximization objective. \\
%Indeed, whenever $X_{c, b, s} \cdot Y_{c, w} = 1$, $Z_{c, b, s}$ can attain a
%maximum value of 1, and whenever $X_{c, b, s} \cdot Y_{c, w} = 0$, at least one
%of $X_{c, b, s}$ or $Y_{c, w}$ must be 0, so $Z_{c, b, s}$ will be constrained
%to attain a maximum value of 0, giving us the correct adjacency maximization
%objective. \\

@@

%Thus, our clinician scheduling problem is a multiple objective optimization
%problem. The most common approach to solving multiple objective optimization
%problems is by optimizing a weighted sum of the normalized objective functions,
%as this guarantees the optimal solution to be Pareto optimal [ref \ref{???}].
%This is the approach we decided to use in our problem, to ensure all three
%objectives are considered when finding a solution. Under the assumption that
%each clinician in $\mathcal{C}$ provides all types of services in
%$\mathcal{S}$, the normalized objectives can be written as follows,
%\begin{align}
%	&\bar{Q}_1(X) = \frac{Q_1(X)}{C \cdot B \cdot S} \tag{Block Requests}
%\label{eqn:norm-obj-block-requests}\\
%	&\bar{Q}_2(Y) = \frac{Q_2(Y)}{C \cdot W} \tag{Weekend Requests}
%\label{eqn:norm-obj-weekend-requests} \\
%	&\bar{Q}_3(Z) = \frac{Q_3(Z)}{C \cdot B \cdot S} \tag{Block-Weekend Adjacency}
%\label{eqn:norm-obj-block-weekend-adj}
%\end{align}
%where we divide each of the original objective functions by the sum of the
%absolute values of its coefficients [ref \ref{???}]. The final weighted
%objective is given by
%\begin{equation}
%	\alpha \bar{Q}_1(X) + \beta \bar{Q}_2(Y) + (1 - \alpha - \beta) \bar{Q}_3(Z)
%\end{equation}
%with $0 \leq \alpha, \beta \leq 1$. \\

### 2019-09-07 paper/paper.results.tex
Add hard line breaks for easier editing

\subsection{Infectious Diseases Division}  %main comment  = I think use past tense in writing
We used clinician time-off requests and minimum/maximum requirements from 2015-2018 as input data for the ILP problem. Table \ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using the software as well as the manually-created schedule for data from 2018. The schedule is color-coded to distinguish between the different clinicians assigned. \\

@@

Figure \ref{fig:runtime-blocks} presents the change in runtime when increasing the number of 2-week blocks $B = \{26, 52, 78, 104\}$ in a department with 10 clinicians offering two services. There does not seem to be a strong effect on runtime when we lengthen the time-horizon of the schedule, and in fact the solver was able to find an optimal schedule for all time horizons within 3 seconds. \\ % 'very reasonable amount of time" is vague. what is reasonable? better to give a range in value.

@@

The effect of increasing the number of services %divisions %do you mean services?
on the runtime of the program is shown in Table \ref{tbl:runtime-services-clinicians-comparison}. We executed the algorithm for $S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in total across all services. For 2 concurrent services, a roster of 30 or more clinicians becomes impractical to schedule, as searching for a solution required over 24 hours. We saw similar issues for a roster of 20 or more clinicians assigned to a division with 3 concurrent services. However, when relaxing the NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3 services, and we were able to generate a schedule with upwards of 50 clinicians in under 1.5 seconds. \\  %what does division mean? do you mean services?

>>>>

\subsection{Infectious Diseases Division}  %main comment  = I think use past
%tense in writing
We used clinician time-off requests and minimum/maximum requirements from
2015-2018 as input data for the ILP problem. Table
\ref{tbl:2018-schedule-comparison} presents the optimal schedule generated using
the software as well as the manually-created schedule for data from 2018. The
schedule is color-coded to distinguish between the different clinicians
assigned. \\

@@

Figure \ref{fig:runtime-blocks} presents the change in runtime when increasing
the number of 2-week blocks $B = \{26, 52, 78, 104\}$ in a department with 10
clinicians offering two services. There does not seem to be a strong effect on
runtime when we lengthen the time-horizon of the schedule, and in fact the
solver was able to find an optimal schedule for all time horizons within 3
seconds. \\ % 'very reasonable amount of time" is vague. what is reasonable?
%better to give a range in value.

@@

The effect of increasing the number of services %divisions %do you mean
%services?
on the runtime of the program is shown in Table
\ref{tbl:runtime-services-clinicians-comparison}. We executed the algorithm for
$S = \{1, 2, 3\}$ total services and $C = \{10, 20, 30, 50\}$ clinicians in
total across all services. For 2 concurrent services, a roster of 30 or more
clinicians becomes impractical to schedule, as searching for a solution required
over 24 hours. We saw similar issues for a roster of 20 or more clinicians
assigned to a division with 3 concurrent services. However, when relaxing the
NCB constraint, we saw a great improvement in runtime for divisions with 2 and 3
services, and we were able to generate a schedule with upwards of 50 clinicians
in under 1.5 seconds. \\  %what does division mean? do you mean services?
